Statically Typed String Sanitation Inside a Python

### 2014-12-02 sanitation-psp14/final.tex
Final PSP14 main paper

\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%\usepackage{enumitem}
\usepackage{enumerate}

@@

\maketitle
\begin{abstract}
Web applications must ultimately command systems like web browsers and database engines using strings. Strings derived from improperly sanitized user input  can thus be a vector for command injection attacks. 
In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language. These types come equipped with common operations like concatenation, substitution and coercion, so they  can be used to implement, in a conventional manner, the portions of a web application or application framework that must directly construct command strings. Simple type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks. We specify this type system in a minimal typed lambda calculus, $\lambda_{RS}$.

To be practical, adopting a specialized type system like this should not require the adoption of a new programming language. Instead, we advocate for extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.  
We support this with two contributions. First, we specify a translation from $\lambda_{RS}$ to a language fragment containing only standard strings and regular expressions. Second, taking Python as a language with these constructs, we implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python being developed by the authors.
\end{abstract}

@@

\section{Introduction}\label{intro}
Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}. They arise because web applications, at their boundaries, control external systems using commands represented as  strings. For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings. When these commands include data derived from user input, care must be taken to ensure that the user cannot  subvert the intended command by carefully crafting the data they send. For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability: 
\begin{lstlisting}[numbers=none]
'SELECT * FROM users WHERE name="' + name + '"'
\end{lstlisting}
If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.

To avoid this problem, the program must \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats safely. Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks. Guaranteeing that user input has already been sanitized  before it is used to construct a command is challenging, especially because  sanitization and command construction may occur in separately developed and maintained components. %is often performed in a different component of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.

We observe that many such sanitization techniques can be understood using \emph{regular languages} \cite{cinderella}. For example, \verb|name| must be a string in the language described by the regular expression \verb!([^"\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity.

In this paper, we present a static type system that tracks the regular language a string belongs to. For example, the output of \verb|sanitize| will be a string in the regular language described by the regular expression above (we identify regular languages by the notation $\mathcal{L}\{r\}$). By leveraging closure and decidability properties of regular languages, the type system tracks the language of a string through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern. This makes it possible to implement sanitation functions -- like the one just described -- in a conventional manner. The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types (e.g. along the interfaces with external systems or components that are not under the control of this type system). %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

@@

Waiting for a language designer to build this feature in is unsatisfying in practice. Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical case studies, but these are  difficult to conduct if developers have no way to use the abstraction in real-world code. We take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.

In Sec. \ref{atlang}, we show how to implement the type system fragment we have specified  using \texttt{atlang}, an extensible static type system implemented as a library inside Python. \texttt{atlang} leverages local type inference to control the semantics of literal forms, so regular string types can be introduced using string literals without any run-time overhead. Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead. This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings in a language that web developers have already widely adopted.

@@

\section{Regular String Types, Minimally}\label{calculus}
\noindent
%This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}. 
This section is organized as follows:
\begin{itemize}
\item Section \ref{sec:rs} describes $\lambda_{RS}$. We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
\item Section \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
\item Section \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result for \ref{sec:rs} is preserved under this translation. 
\end{itemize}

The accompanying technical report \cite{psptr} contains more detailed proofs, further discussion of string substitution and language replacement, and
a single sheet containing Figures \ref{fig:glambdas}-\ref{fig:tr} for the reader's convenience.

\subsection{The Language of Regular Strings}\label{sec:rs}
In this section, we define a minimal typed lambda calculus with regular string types called $\lambda_{RS}$. The syntax of $\lambda_{RS}$ is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics in Figure \ref{fig:dlambdas}.

There are two type constructors in $\lambdas$, $\rightarrow$ and $\textsf{stringin}$. Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}. Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression. Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e. $s \in \Sigma^\star$, defined in the usual way). The rule \textsc{S-T-Stringin-I} requires that $s \in \lang{r}$.
 %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.

$\lambdas$ provides several familiar operations on strings. The type system  relates these operations over strings to corresponding operations over the  regular languages they belong to. 
Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for static analysis of sanitation protocols (we will see a complete example in Section \ref{atlang}).

@@

\subsubsection{Concatenation and String Decomposition}

@@

Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings. Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language. The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables. The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics. This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).

The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.

@@

\begin{defn}[Definition of $\lhead{r}$]
$\lhead{r}$ is defined in terms of the structure of $r$:
{\footnotesize
\begin{align*}
\lhead{r}                 &= \lhead{r, \epsilon} \\                             
\lhead{\epsilon, r'}      &= \epsilon \\                                        
\lhead{a, r'}             &= a \\                                               
\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\                                
\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\               
\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}         
\end{align*}} 
\end{defn}
We use a two-argument auxiliary $\lhead{r, r'}$ in the definition of $\lhead{r}$ because when $r = q^*\cdot t$,
$\lhead{r}$ needs to ``remember" $t$ in case $q$ is iterated zero times.

@@

The rule 
 \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$. Language inclusion is decidable. As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any checks. The rule 
\textsc{S-T-Check}, conversely, does not perform any static checks on the two languages involved, only checking that the two branches have the same type. The checks are performed dynamically, by rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.

In our calculus, both forms of coercion are explicit. For safe coercions, it is often useful for the coercion to be performed implicitly. This can be seen as a form of subtyping for regular string types \cite{gunter, fulton13, xhaskell2}. 
In practice, subtyping is crucial to the usability of our system due to the nature of the replacement operator.
For simplicity, we do not include this operator; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton13}. The implementation discussed in Section \ref{atlang} also provides subtyping between regular string types.

\subsubsection{Replacement}\label{sec:replace}
The premier operation for working with regular strings in $\lambdas$ is  replacement, written $\sisubst{r}{e_1}{e_2}$. 
It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given. The evaluation rule \textsc{S-E-replace} is defined in terms of $\lsubst{r}{s_1}{s_2}$. For simplicity, we do not define replacement in this paper; see \cite{psptr}.

\begin{defn}[$\tt{subst}$]
$\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
\end{defn}

The typing rule \textsc{S-T-Replace} requires computing the regular language of the string resulting from this operation given the languages of $s_1$ and $s_2$, written as $\lreplace{r}{r_1}{r_2}={r'}$. 
Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata.

A complete definition of ${\sf lreplace}$ would also give an rewrite system with correctness and termination proofs, which is beyond the scope of this paper. Instead, we provide an abstract definition of the operation and state necessary properties.

\begin{defn}[$\tt{lreplace}$]
  $\lreplace{r}{r_1}{r_2}$ relates $r, r_1,$ and $r_2$ to
  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty). This procedure saturates the string.
\end{defn}

\begin{prop}[Closure] \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{prop}
\begin{prf}
Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
equals $x$; $x_{post}$ can be computed in a similar way after reversal.
Then $r'$ is $x_{pre}r_2x_{post}$.
\end{prf}

\begin{prop}[Replacement Correspondence] \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then\\ $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{prop}

In this section, we establish type soundness for $\lambdas$.
The theorem relies upon the definitions and propositions given above and some basic properties of regular languages.

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}
\item 
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
\item 
For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\item 
Regular languages are closed under reversal.
\end{enumerate}
\end{lem}

@@

We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
\begin{lem}
  If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}

Safety for the string fragment of $\lambdas$ requires validating that the type system's definition is justified
by our theorems about regular languages.

We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
The simply typed lambda calculus terminates, and our conservative extension clearly
terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety would hold, but a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.

\begin{thm}[Type Safety.] \label{thm:typesafety}
  If $\emptyset \vdash e : \sigma$ 
  then $e \sreduces v$ and $\emptyset \vdash v : \sigma$.
\end{thm}
\begin{prf}
By induction on the typing relation.
The \textsc{S-T-Concat} case requires Lemma \ref{thm:regexprops}.1, the \textsc{S-T-Check} case requires Lemma \ref{thm:regexprops}.2 
and the S-T-Replace case appeals to Proposition \ref{thm:substcorrespondence}.
\end{prf}

@@

\begin{lem}[Normalization for String Fragment of $\lambdas$]\label{thm:cfs}
  If $\emptyset \vdash e : \stringin{r}$ and $e \sreduces v$ then $v=\rstr{s}$. 
\end{lem}

\subsubsection{The Security Theorem}\label{sec:securitythm}

The chief benefit of $\lambdas$ is its security theorem, which states that any value of
a regular string type is recognized by the regular language corresponding to the regular expression the type is indexed by.
This is beneficial because this ensure that membership in a regular language known to be secure becomes manifest in the type of the string, rather than being a property that must be established extralinguistically.

\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
  If  $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
\end{thm}
\begin{prf}
  The theorem follows directly from type safety, canonical forms for $\lambdas$, and inversion of \textsc{S-T-Stringin-I}, the typing rule for terms of the form $\rstr{s}$.
\end{prf}

@@

<$\iota$> ::= $\dot{v}$ | $x$ | $\iota(\iota)$ \hfill target terms \alt
   $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
   $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$
\end{grammar}
\vspace{-5px}
\caption{Syntax of the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}
Our next major technical result, stated in Sec. \ref{sec:tr}, establishes that the security property
is preserved under translation into $\lambdap$, which we now define.

@@

\inferrule[P-T-Check]
{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
\end{mathpar}
\vspace{-10pt}
\caption{Typing rules for $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:slambdap}
\end{figure}

>>>>

\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
%\usepackage{enumitem}
\usepackage{enumerate}%

@@

\maketitle
\begin{abstract}
Web applications must ultimately command systems like web browsers and database engines using strings. Strings derived from improperly sanitized user input  can as a result be a vector for command injection attacks. 
In this paper, we introduce \emph{regular string types}, which classify strings constrained statically to be in a regular language specified by a regular expression. Regular strings support standard string operations like concatenation and substitution, as well as safe coercions, so they  can be used to implement, in an essentially conventional manner, the pieces of a web application or framework that  handle strings arising from user input. Simple type annotations at function interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks. We specify this type system first as a minimal typed lambda calculus, $\lambda_{RS}$.

To be practical, adopting a specialized type system like this should not require the adoption of a new programming language. Instead, we advocate for extensible type systems: new type system fragments like this should be implemented as libraries atop a mechanism that guarantees that they can be safely composed.  
We support this with two contributions. First, we specify a translation from $\lambda_{RS}$ to a calculus with only standard strings and regular expressions. Then, taking Python as a language with these constructs, we implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python.
\end{abstract}

@@

\section{Introduction}\label{intro}
Command injection vulnerabilities are among the most common and severe security vulnerabilities in modern web applications \cite{OWASP}. They arise because web applications, at their boundaries, control external systems using commands represented as  strings. For example, web browsers are controlled using HTML and Javascript sent from a server as a string, and database engines execute SQL queries also sent as strings. When these commands include substrings derived from user input, care must be taken to ensure that the user cannot  subvert the intended command. For example, a  SQL query constructed using string concatenation exposes a SQL injection vulnerability: 
\begin{lstlisting}[numbers=none]
'SELECT * FROM users WHERE name="' + name + '"'
\end{lstlisting}\vspace{-3px}
If a malicious user enters the name \lstinline{'"; DROP TABLE users --'}, the entire database could be erased.

To avoid this problem, developers are cautioned to \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that escapes double quotes and existing backslashes   with a backslash, which SQL treats securely. Alternatively, it might HTML-encode special characters, which would head off both SQL injection attacks and cross-site scripting attacks. \emph{Guaranteeing} that user input has already been sanitized  before it is used to construct a command is challenging, especially because  sanitization and command construction may occur in separately developed and maintained components. %is often performed in a different component of a program than where commands are ultimately constructed.% Note, for example, that \verb|sanitize| is not idempotent, so it should only be called once.

We observe that many such sanitization techniques can be specified using \emph{regular languages} \cite{cinderella}. For example, for the command constructed above to be secure, \verb|name| must be a string in the language described by the regular expression \verb!([^"\\]|(\")|(\\))*! -- a sequence of characters other than quotation marks and backslashes; these can only appear escaped. This conventional syntax for regular expression patterns can be understood to desugar, in a standard way, to the syntax for regular expressions shown in Figure \ref{fig:regex}, where $r \cdot r$ is sequencing and $r + r$ is  disjunction. We will work with this ``core'' for simplicity in the remainder.

In this paper, we present a static type system that tracks the regular language a string belongs to (we identify regular languages by the notation $\mathcal{L}\{r\}$). For example, the output of \verb|sanitize| will not simply be \verb|string|, but rather \verb|stringin[r]|, where \verb|r| is the regular expression above. By leveraging closure and decidability properties of regular languages, the type system will be able to closely track the language a string belongs to through uses of a number of operations, including \emph{replacement} of substrings matching a given pattern. This makes it possible to implement sanitization functions -- like the one just described -- in a conventional manner. The result is a system where the fact that a string has been {correctly} sanitized becomes manifest in its type. Missing calls to sanitization functions can thus be detected statically, and, crucially, so can \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types. %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

@@

Waiting for a language designer to build this feature in is unsatisfying in practice. %Moreover, we also face a ``chicken-and-egg problem'': justifying its inclusion into a commonly used language benefits from empirical case studies, but these are  difficult to conduct if developers have no way to use the abstraction in real-world code. 
We take the position that a better path forward for the community is to work within a programming language where such type system fragments can be introduced modularly and orthogonally, as libraries.

In Sec. \ref{atlang}, we show how to implement the type system and translation from Sec. \ref{calculus} using \texttt{atlang}, an extensible static type system implemented as a library inside Python. \texttt{atlang} leverages local type inference to control the semantics of literal forms, so regular string types can be introduced using string literals without any run-time overhead. Coercions that are known to be  safe due to a sublanguage relationship are performed implicitly, also without run-time overhead. This results in a \emph{usably secure} system: working with regular strings differs little from working with standard strings in a language that web developers have already widely adopted.

@@

\section{Regular String Types, Minimally}\label{calculus}
\noindent
%This will serve as the source language for our translation to a calculus with only standard strings and regular expressions, defined in Section~\ref{sec:tr}. 
This section is organized as follows:
\begin{itemize}
\item Sec. \ref{sec:rs} describes $\lambda_{RS}$ and its metatheory.% We also give proof outlines of  type safety and correctness theorems and relevant propositions about regular languages.
\item Sec. \ref{sec:p} describes a simple target language, $\lambdap$, with a minimal regular expression library. In Section \ref{atlang}, we will take Python to be such a language.
\item Sec. \ref{sec:tr} describes the translation from $\lambdas$ to $\lambdap$ and ensures the correctness result from Sec. \ref{sec:rs} is preserved under this translation. 
\end{itemize}

The accompanying technical report \cite{psptr} contains more detailed proofs and further discussion of  design choices.% Figures \ref{fig:glambdas}-\ref{fig:tr} for the reader's convenience.

\subsection{The Language of Regular Strings}\label{sec:rs}
In this section, we define a typed lambda calculus with regular string types called $\lambda_{RS}$. Its syntax is specified in Figure \ref{fig:glambdas}, its static semantics in Figure \ref{fig:slambdas} and its evaluation semantics, given here in big-step style, in Figure \ref{fig:dlambdas}.

There are two type constructors in $\lambdas$: $\rightarrow$ and $\textsf{stringin}$. Arrow types classify functions, which are introduced via lambda abstraction, $\lambda x.e$, and can be applied, written $e(e)$, in the usual way \cite{pfpl}. Regular string types are of the form $\stringin{r}$, where $r$ is a regular expression. Values of such regular string types take the form $\strin{s}$, where $s$ is a string (i.e. $s \in \Sigma^\star$, where the Kleene star is defined in the usual way). The rule \textsc{S-T-Stringin-I} requires that $s \in \lang{r}$.
 %The remaining operations on terms of type $\stringin{r}$ preserve this property. %The remaining operators operate on regular strings.

$\lambdas$ provides several familiar operations on strings. The type system  relates these operations over strings to corresponding operations over the  regular languages they belong to. 
Since these operations over regular languages are known to be closed and decidable, we can use these operations as a basis for implementing and reasoning about sanitation protocols, as we will discuss below.

@@

\subsubsection{Concatenation}

@@

\subsubsection{String Decomposition}
Whereas concatenation allows the construction of large strings from smaller strings,  $\strcase{e}{e_0}{x,y.e_1}$  allows the decomposition, or elimination, of large strings into smaller strings. Intuitively, this operation branches based on whether a string is empty or not, exactly analagous to case analysis on lists in a functional language. The branch for a non-empty string ``peels off" the first character, binding it and the remainder of the string to specified variables. The evaluation rules \textsc{S-E-Case-$\epsilon$} and \textsc{S-E-Case-Concat} express this semantics. This construct can be used to implement a standard string indexing operation as a function, given a suitable definition of natural numbers (omitted).

The typing rule \textsc{S-T-Case} must determine a suitable type for the head and tail of the string. The regular expression recognizing any one-character prefix of the strings in $\lang{r}$ is easily defined.

@@

\begin{defn}[Definition of $\lhead{r}$] 
{\footnotesize
\begin{align*}
\lhead{r}                 &= \lhead{r, \epsilon}
\end{align*}}We use a two-argument auxiliary $\lhead{r, r'}$ in the definition of $\lhead{r}$ because when $r = q^*\cdot r'$,
$\lhead{r}$ needs to ``remember" $r'$ in case $q$ is iterated zero times:
{\footnotesize
\begin{align*}
\lhead{\epsilon, r'}      &= \epsilon \\                                        
\lhead{a, r'}             &= a \\                                               
\lhead{r_1\cdot r_2, r'}  &= \lhead{r_1, r_2} \\                                
\lhead{r_1 + r_2, r'}     &= \lhead{r_1, r'} + \lhead{r_2, r'} \\               
\lhead{r^*, r'}           &= \lhead{r', \epsilon} + \lhead{r, \epsilon}         
\end{align*}} 
\end{defn}

@@

The rule 
 \textsc{S-T-SafeCoerce} checks for language inclusion, which we write $\lang{r_1} \subseteq \lang{r_2}$. Language inclusion is decidable. As a result, the rule \textsc{S-E-SafeCoerce} does not need to perform any dynamic checks. The rule 
\textsc{S-T-Check}, conversely, does not perform any static checks on the two languages involved, only checking that the two branches have the same type. The checks are performed dynamically, by rules \textsc{S-E-Check-Ok} and \textsc{S-E-Check-NotOk}.

In our calculus, both forms of coercion are explicit. For safe coercions, it is often useful for the coercion to be performed implicitly. This can be seen as a form of subtyping for regular string types \cite{gunter, fulton13, xhaskell2}. 
In practice, subtyping will be crucial to the usability of our system due to the nature of the replacement operator.
For simplicity, we do not specify subtyping in our core calculus; however, subtyping for regular strings is present in previous treatments of our work \cite{fulton13}. The implementation discussed in Section \ref{atlang} also provides subtyping between regular string types.

\subsubsection{Replacement}\label{sec:replace}
The premier operation for working with regular strings in $\lambdas$ is  replacement, written $\sisubst{r}{e_1}{e_2}$. 
It behaves analogously to \lstinline{str_replace} in PHP and \lstinline{String.replace} in Java, differing in that the replacement pattern $r$ must be statically given. The evaluation rule \textsc{S-E-replace} is defined in terms of the metafunction $\lsubst{r}{s_1}{s_2}$.

\begin{defn}[$\sf{replace}$]
$\lsubst{r}{s_1}{s_2} = s$ such that all substrings of $s_1$ in $\lang{r}$ are replaced with $s_2$.
\end{defn}
The typing rule \textsc{S-T-Replace} thus requires computing the regular language of the string resulting from this operation given knowledge of the languages of $s_1$ and $s_2$, written as $\lreplace{r}{r_1}{r_2}={r'}$.

Given an automata-oriented interpretation of regular languages, it may be helpful to think in terms of replacing sub-automata. A complete definition of ${\sf lreplace}$ would consist of a rewrite system based on this intuition, with correctness and termination proofs, which is beyond the scope of this paper. Instead, we provide an abstract definition of the operation and state necessary properties.

\begin{defn}[$\sf{lreplace}$]
  $\lreplace{r}{r_1}{r_2}$ relates $r, r_1,$ and $r_2$ to
  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty). This procedure saturates the string.
\end{defn}

% \begin{prop}[Closure] \label{thm:total}
%   If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular languages, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
% \end{prop}
% \begin{prf}
% Algorithms for the inclusion problem may be adopted to identify any sublanguage $x \subseteq r$ of $r_1$.
% The language $x_{pre}$ can be computed by taking total derivatives until the remaining language
% equals $x$; $x_{post}$ can be computed in a similar way after reversal.
% Then $r'$ is $x_{pre}r_2x_{post}$.
% \end{prf}

\begin{prop}[Replacement Correspondence] \label{thm:substcorrespondence}
  Given strings $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ we have $$\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$$
\end{prop}

\subsubsection{Metatheory of $\lambdas$}

In this section, we establish some basic metatheoretic properties of $\lambdas$.
These rely upon the definitions and propositions given above and some basic properties of regular languages.

\begin{lem}[Properties of Regular Languages.] \label{thm:regexprops}
~
\begin{enumerate}
\item 
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1\cdot r_2}$.
\item 
For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
%\item 
%Regular languages are closed under reversal.
\end{enumerate}
\end{lem}

@@

% We also need to establish a well-formedness lemma, which simply expresses the totality of the operations over regular expressions used as premises in the rules in Figure \ref{fig:slambdas}. 
% \begin{lem}
%   If $\sctx \vdash e : \stringin{r}$ then $r$ is a well-formed regular expression.
% \end{lem}

Type preservation for $\lambdas$ requires validating that the statics are consistent with the dynamics. We give a full proof of type safety for a variant of the calculus with a small step semantics in \cite{psptr}, but for concision, it is more straightforward to explain the semantics with a big step semantics here.

% We avoid the most significant issues inherent to proofs related to big-step semantics by avoiding non-termination in our specification.
% The simply typed lambda calculus terminates, and our conservative extension clearly
% terminates because new binding structures are not introduced and termination (i.e. decidability) of ${\sf subst}$ and ${\sf lreplace}$ is known. Even in a non-terminating calculus, type safety would hold, but a more careful treatment or treatment for a language with non-termination would need to proceed by either a coinductive argument \cite{coinductionoo} or  with a small-step semantics.

\begin{thm}[Type Preservation.] \label{thm:typesafety}
  If $\emptyset \vdash e : \sigma$ 
  and $e \sreduces v$ then $\emptyset \vdash v : \sigma$.
\end{thm}
\begin{prf}
By induction on the typing relation.
The \textsc{S-T-Concat} case requires Lemma \ref{thm:regexprops}.1, the \textsc{S-T-Check} case requires Lemma \ref{thm:regexprops}.2 
and the S-T-Replace case appeals to Proposition \ref{thm:substcorrespondence}.
\end{prf}

@@

\begin{lem}[Canonical Forms for Regular Strings]\label{thm:cfs}
  If $\emptyset \vdash v : \stringin{r}$ then $v=\rstr{s}$ and $s \in \lang{r}$. 
\end{lem}
\begin{prf}
The only typing rule that applies is \textsc{S-T-Stringin-I}. The conclusion is the first premise.
\end{prf}
\subsubsection{The Security Theorem}\label{sec:securitythm}

The chief benefit of $\lambdas$ is its security theorem, which states that any term of
a regular string type that evaluates to a value will evaluate to a regular string recognized by the regular language corresponding to the regular expression the type is indexed by.
This is beneficial because this ensure that membership in a regular language known to be secure becomes manifest in the type of the string, rather than being a property that must be established extralinguistically.

\begin{thm}[Correctness of Input Sanitation for $\lambdas$]\label{thm:scorrect}
  If  $\emptyset \vdash e : \stringin{r}$ and $e \sreduces \rstr{s}$ then $s \in \lang{r}$.
\end{thm}
\begin{prf}
  The theorem follows directly from type preservation and canonical forms above.
\end{prf}

@@

<$\iota$> ::= $\dot{v}$ | $x$ | $\iota(\iota)$ \hfill target terms \alt
   $\tconcat{\iota}{\iota}$ | $\pstrcase{\iota}{\iota}{x,y.\iota}$ \alt
   $\tcheck{\iota}{\iota}{\iota}{\iota}$ | $\preplace{\iota}{\iota}{\iota}$
\end{grammar}
\vspace{-5px}
\caption{Syntax of the target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}
Our next major technical result, stated in Sec. \ref{sec:tr}, establishes that the security property
is preserved under translation into $\lambdap$. The system $\lambdap$ is another straight-forward extension of a simply typed lambda calculus
with a $\str$ type and a $\regex$ type, as well as 
some operations -- such as concatenation and replacement -- found in the standard libraries of many programming languages.
The operations of $\lambdap $ correspond to ``run-time'' versions of the operations performed statically in $\lambdas$ in a way made precise by the translation rules described in the next section. 
%Unlike $\lambdas$, $\lambdap$ does not statically track the effects of string operations.
The language $\lambdap$ is so-called because it is reminscent of popular web programming languages,
such as \textbf{P}ython or \textbf{P}HP, albeit statically typed. We will discuss an implementation within the former in the next section.

@@

\inferrule[P-T-Check]
{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
{\tctx \vdash \tcheck{\iota_r}{\iota_1}{\iota_2}{\iota_3} : \sigma}
\end{mathpar}
\vspace{-10pt}
\caption{Typing rules for $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:slambdap}
\end{figure}
Type preservation for $\lambdap$ is essentially trivial, but is necessary in order to establish the 
correctness of our translation.
Again, we give a small step semantics and address type safety more completely in \cite{psptr}.

\begin{thm}[Safety for $\lambdap$] If $\emptyset \vdash \iota : \tau$ 
  and $\iota \sreduces \dot v$ then $\emptyset \vdash \dot v : \tau$.
\end{thm}
